Comparison of dimensionality reduction schemes for parallel global optimization algorithms \thanks{The study was supported by the Russian Science Foundation, project No 16-11- 10150}

### 2018-04-13 2018/RSD/MultiEvolvents/RSD.tex
RSD: english version

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
%%%%%ДОБАВИЛ ДЛЯ РУССКОГО ТЕКСТА
\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}
%%%%%
\usepackage{eqnarray,amsmath}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\setlength{\textfloatsep}{4pt plus 1.0pt minus 1.5pt}
\setlength{\floatsep}{4pt plus 1.0pt minus 1.5pt}

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
%
\title{Comparison of dimensionality reduction schemes for parallel global optimization algorithms
\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov \and
Vladislav Sovrasov \and
Ilya Lebedev}
%
\authorrunning{K. Barkalov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia
\email{konstantin.barkalov@itmm.unn.ru}\\
\email{sovrasov.vlad@gmail.com}\\
\email{ilya.lebedev@itmm.unn.ru}\\
\url{http://hpc-education.unn.ru/основные-напраления/глобальная-оптимизация} }
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This work considers a parallel algorithms for solving multi-extremal optimization problems. Algorithms are developed within the framework of the information-statistical approach and implemented in a parallel solver “Globalizer” . The optimization problem is solved by reducing the multidimensional problem to a set of joint one-dimensional problems that are solved in parallel. Five types of Peano-type space-filling curves are employed to reduce dimension. The results of computational experiments carried out on several hundred test problems are discussed.

@@

%------------------------------------------------------------------------------
\section{Introduction}\label{sec:intro}

%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer\cite{globalizerSystem}, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\) space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .

@@

\par
The computational scheme obtained as a result of the dimensionality reduction consists of the following:
\begin{itemize}
  \item The optimization algorithm performs the minimization of the reduced one-dimensional
  function \(\varphi(y(x))\) from (\ref{eq:oneDimTask}),
  \item After determining the next trial point \(x\), a multidimensional image \(y\) is calculated by using the
mapping \(y(x)\),
  \item The value of the initial multidimensional function \(\varphi(y)\) is calculated at the point \(y\in D\),
  \item The calculated value \(z=\varphi(y)\) is used further as the value of the reduced one-dimensional function \(\varphi(y(x))\) at the point \(x\).
\end{itemize}

@@

\begin{figure}[ht]
    \centering
    \subfloat[Two mappings]{{\includegraphics[width=.6\textwidth]{pictures/shifted.pdf}}\label{fig:shifted_ev}}
    \subfloat[Hypercubes $D_l$]{{\includegraphics[width=.4\textwidth]{pictures/shifted_cube.png}}\label{fig:shifted_cube}}
    \caption{Multiple mappings}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.6\linewidth]{pictures/rotated_1.pdf}
  \caption{Two rotated evolvents on the same plane}
  \label{6_fig_9}
\end{figure}

%------------------------------------------------------------------------------
\subsection{Non injective evolvent}

Как уже было сказано в секции \ref{sec:shifted}, потеря информации о близости точек в многомерном пространстве может быть частично скомпенсирована использованием множественных отображений $Y_L(x)=\{y^1(x),...,y^L(x)\}$. Однако, сама по себе кривая типа Пеано сохраняет в себе часть этой информации: она не является инъективым отображением, поэтому имея один образ $y(x)\in \mathbb{R}^N$, можно получить несколько несколько отличных $x$ прообразов $t_j\in[0,1], t_j \not = x$, которые затем могут быть добавлены в поисковую информацию индексного метода.

Кривая типа Пеано, используемая в (\ref{eq:oneDimTask}) для редукции размерности, определяется через предельный переход,
поэтому не может быть вычислена непосредтвенно. При численной оптимизации используется некоторое её приближение, являющееся
инъективной кусочно-линейной кривой. В \cite{strongin1978} было предложено неинъективное отображение равномерной сетки на отрезка $[0,1]$ на равномерную сетку в гиперкубе $D$. Каждый многомерный узел может иметь до $2^N$ одномерных прообразов. На рис. \ref{fig:noninjective} изображена кривая с самопересечениями, полученная при соединении узлов грубой многомерной сетки в порядке следования их прообразов из отрезка $[0,1]$, а также отмечана точка, имеющая 3 прообраза.

Недостатком неинъективной развёртки является потенциально большое количетво прообразов (до $2^N$) и невозможность использования параллельной схемы для множественных отображений из секции \ref{sec:parallel_evolvents}.

@@

\begin{figure}[ht]
    \centering
    \subfloat[Smooth evolvent]{{\includegraphics[width=.6\textwidth]{pictures/smooth.pdf}}\label{fig:smooth}}
    \subfloat[Numerical non injective evolvent]{{\includegraphics[width=.4\textwidth]{pictures/noninjective.pdf}}\label{fig:noninjective}}
    \caption{Different evolvents}
\end{figure}

@@

%------------------------------------------------------------------------------
\subsection{Сравнение последовательных разверток}
\label{sec:seq_comp}
С целью понять, обладает ли какой-либо из перечисленных ранее типов развёрток существенным преимуществом над другими, были построены операционные характеристики индексного метода с различными типами развёрток на классах GKLS 2d Simple и GKLS 3d Simple. The global minimum was considered to be found if the algorithm generates a trial point $y^k$ in the $\delta$-vicinity of the global minimizer, i.e. $\left\|y^k-y^\ast\right\|_\infty\leq\delta$. The size of the vicinity was selected as $\delta = 0.01\left\|b-a\right\|_\infty$. In case of GKLS $\delta=0.01$.

Во всех экспериментах параметр плотности построения развёрток $m=12$. Минимальное значение пераметра надёжности \(r\) было найдено для каждого типа развёртки перебором по равномерной сетке с шагом \(0.1\).

На классе GKLS 2d Simple при минимальном \(r\) неинъективная и гладкая развёртка обеспечивают более быструю сходимость (рис. \ref{fig:gkls2d_opt}). То же самое, наблюдается и при \(r=5.0\) (рис. \ref{fig:gkls2d_acc}). В последнем случае сдвиговая и вращаемая развёртки начинают отставать от остальных, т.к. значение \(r=5.0\) является завышенным для них.
\begin{figure}[ht]
    \centering
    \subfloat[$r=5.0$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_same_r_opt_pt_op.pdf}}\label{fig:gkls2d_acc}}
    \subfloat[Minimal $r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_opt_pt_op.pdf}}\label{fig:gkls2d_opt}}
    \caption{Operating characteristics on GKLS 2d Simple class}
\end{figure}

На классе GKLS 2d Simple при минимальном \(r\) неинъективная и множественные развёртки имеют значительное преимущество над единственной развёрткой (рис. \ref{fig:gkls3d_opt}). Значение \(r=4.5\) является завышенным для вращаемой и сдвиговой развёрток (рис. \ref{fig:gkls3d_acc}).

\begin{figure}[ht]
    \centering
    \subfloat[$r=4.5$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_same_r_opt_pt_op.pdf}}\label{fig:gkls3d_acc}}
    \subfloat[Minimal $r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_opt_pt_op.pdf}}\label{fig:gkls3d_opt}}
    \caption{Operating characteristics on GKLS 3d Simple class}
\end{figure}

\paragraph{Накладные расходы при использовании сдвиговой развёртки.}
Во всех представленных выше экспериментах при построении операционной характеристики учитывалось количество вычислений целевой функции из класса GKLS, однако в случае сдвиговой развёртки индексный метод решает задачу с ограничением \(g_0\) из (\ref{6_g0}). В точках, где \(g_0\) нарушено, значение целевой функции не вычисляется. Эти точки, тем не менее, хранятся в поисковой информации, создавая дополнительные расходы вычислительных ресурсов. В таблице \ref{tab:shifted_g0} приведено среднее количество обращение к \(g_0\) и целевой функции. При \(L=3\) ограниение \(g_0\) вычисляется почти в 20 раз чаще, чем целевая функция \(\varphi\), т. е. \(95\%\) всей поисковой информации приходится на вспомогательные точки. Такие накладные расходы приемлемы при решении задач малой размерности с трудоёмкими целевыми функциями, но при росте размерности и общего количества испытаний выгоднее использовать другие типы развёрток.

\begin{table}
\begin{center}
\caption{Среднее количество вычислений \(g_0\) и \(\varphi\) при решении задач класса GKLS 3d Simple с помощью сдвиговой развёртки}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
  $L$ & $calc(g_0)$ & $calc(\varphi)$ & $\frac{calc(g_0)}{calc(\varphi)}$ ratio \\
  \hline
  2 & 96247.9  & 6840.14 & 14.07\\
  \hline
  3 & 153131.0 & 7702.82 & 19.88\\
  \hline
  \end{tabular}
  \label{tab:shifted_g0}
\end{center}
\end{table}

%------------------------------------------------------------------------------
\subsection{Параллельные вращаемые развертки}
\label{sec:results_parallel}
Чтобы оценить эффективность параллельного алгортима из секции \ref{sec:parallel_evolvents} были проведены вычислительные эксперименты на классах GKLS 4d (Hard, Simple) и GKLS 5d (Hard, Simple). Значения \(r\) во всех экспериментах равно 5.0, размер \(\delta\)-окрестности известного решения увеличен до 0.3. При решении серий задач использовалось до 8 узлов кластера и до 32 вычислительных потоков на каждом узле.

В таблице \ref{tab:iterations} приведено среднее количество итераций при решении 100 задач из каждого из рассматриваемых классов. При увеличении числа узлов и числа потоков на каждом узле количество итераций заметно сокращается (за исключением класса GKLS 4d Simple при переходе с 1 узла на 4 узла в однопоточном режиме).

@@

Если считать, что затраты на параллелизм пренебрежимо малы по сравнению с затратами на вычисление целевых функций в задачах оптимизации, то ускорение по времени от использования параллельного метода будет равно ускорению по итерациям. Однако, в действительности это предположение не всегда справедливо. Во всех численных экспериментах время вычисления целевой функции занимает примерно $10^{-3}$с. В таблице \ref{tab:speedup} приведено ускорение по итерациям и в куглых скобках ускорение по времени. В первой строке таблицы, соответствующей последовательному режиму, в скобках приведено среднее время решения одной задачи. Из таблицы видно, что для класов GKLS 4d выгоднее использовать один узел в многопоточном режиме, тогда как для решения более сложных пятимерных задач лучше использовать несколько узлов, каждый из которых работает в параллельном режиме.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%%%%%
\usepackage{eqnarray,amsmath}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\setlength{\textfloatsep}{4pt plus 1.0pt minus 1.5pt}
\setlength{\floatsep}{4pt plus 1.0pt minus 1.5pt}

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
%
\title{Comparison of dimensionality reduction schemes for parallel global optimization
algorithms
\thanks{The study was supported by the Russian Science Foundation, project No 16-11-
10150}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov \and
Vladislav Sovrasov \and
Ilya Lebedev}
%
\authorrunning{K. Barkalov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia
\email{konstantin.barkalov@itmm.unn.ru}\\
\email{sovrasov.vlad@gmail.com}\\
\email{ilya.lebedev@itmm.unn.ru}\\
\url{http://hpc-education.unn.ru/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This work considers a parallel algorithms for solving multi-extremal optimization problems.
Algorithms are developed within the framework of the information-statistical approach and
implemented in a parallel solver Globalizer. The optimization problem is solved by reducing
the multidimensional problem to a set of joint one-dimensional problems that are solved in
parallel. Five types of Peano-type space-filling curves are employed to reduce dimension. The
results of computational experiments carried out on several hundred test problems are discussed.

@@

%------------------------------------------------------------------------------
\section{Introduction}\label{sec:intro}
%\begin{Russian}

In the present paper, the parallel algorithms for solving the multiextremal optimization problems
are considered. In the multiextremal problems, the opportunity of reliable estimate of the global
optimum is based principally on the availability of some information on the function known
{\textit a priori} allowing relating the probable values of the optimized function to the known
values at the points of performed trials. Very often, such an information on the problem being
solved is represented in the form of suggestion that the objective function $\varphi(y)$ satisfies
Lipschitz condition with the constant $L$ not known {\textit a priori} (see, for example,
\cite{Jones,Gablonsky,Evtushenko}). At that, the objective function could be defined by a
program code i. e. could represent a black-box-function. Such problems are presented in the
applications widely (problems of optimal design of objects and technological processes in
various fields of technology, problems of model fitting according to observed data in scientific
research, etc.).

Many methods destined to solving the problems of the class specified above reduce the solving
of a multidimensional problem to solving the one-dimensional subproblems implicitly (see, for
example, the methods of diagonal partitions \cite{Sergeyev2006,SergeyevKvasov2015} or
simplicial partitions \cite{Zilinskas2008,Zilinskas2014}). In the present work, we will use the
approach developed in Lobachevskii State University of Nizhny Novgorod based on the idea of
the dimensionality reduction with the use of Peano space-filling curves $y(x)$ mapping the
interval $[0,1]$ of the real axis onto an $n$-dimensional cube continuously and unambiguously.

Several methods of constructing the evolvents approximating the theoretical Peano curve have
been proposed In Refs. \cite{strongin1978,Strongin1992,Gergel2009,Goryachih2017}. These
methods were implemented in the Globalizer software system \cite{globalizerSystem}. The goal
of the present study was comparing the properties of the evolvents and the selecting the most
suitable ones for the use in the parallel global optimization algorithms.

%\end{Russian}
%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function
\(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\)
space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .

@@

\par
The computational scheme obtained as a result of the dimensionality reduction consists of the
following:
\begin{itemize}
  \item The optimization algorithm performs the minimization of the reduced one-dimensional
  function \(\varphi(y(x))\) from (\ref{eq:oneDimTask}),
  \item After determining the next trial point \(x\), a multidimensional image \(y\) is calculated by
using the mapping \(y(x)\),
  \item The value of the initial multidimensional function \(\varphi(y)\) is calculated at the point
\(y\in D\),
  \item The calculated value \(z=\varphi(y)\) is used further as the value of the reduced one-
dimensional function \(\varphi(y(x))\) at the point \(x\).
\end{itemize}

@@

\begin{figure}[ht]
    \centering
    \subfloat[Two
mappings]{{\includegraphics[width=.6\textwidth]{pictures/shifted.pdf}}\label{fig:shifted_ev}
}
    \subfloat[Hypercubes
$D_l$]{{\includegraphics[width=.4\textwidth]{pictures/shifted_cube.png}}\label{fig:shifted_c
ube}}
    \caption{Multiple mappings}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.6\linewidth]{pictures/rotated.pdf}
  \caption{Two rotated evolvents on the same plane}
  \label{6_fig_9}
\end{figure}

%------------------------------------------------------------------------------
\subsection{Non-Univalent evolvent}
%\begin{Russian}

As it has been already mentioned above (Sec.~\ref{sec:shifted}), the loss of information on the
proximity of the points in the multidimensional space could be compensated in part by the use
of multiple mappings $Y_L(x)=\{y^1(x),...,y^L(x)\}$. However, the Peano-type curve preserves
a part of this information itself: it is not an injective mapping. Therefore, if a single image
$y(x)\in \mathbb{R}^N$ is available, on can obtain several different $x$ preimages
$t_j\in[0,1], t_j \not = x$, which could be added into the search information of the index
method later.

The Peano-type curve used in (\ref{eq:oneDimTask}) for the dimensionality reduction is
defined via the transition to the limit. Therefore, it cannot be computed directly. In the
numerical optimization, some approximation of this curve is used, and it is an injective piecewise-
linear curve. In Ref.~\cite{strongin1978}, a non-univalent mapping of a uniform grid in the
interval $[0,1]$ onto a uniform grid in a hypercube $D$ has been proposed. Each
multidimensional node can have up to $2^N$ one-dimensional preimages. In
Fig.~\ref{fig:noninjective}, the grid in the $\mathbb{R}^2$ space is marked by the crosses, for
two nodes of which the corresponding one-dimensional preimages from $[0,1]$ are pointed
(marked by the squares and circles). Each node mentioned above has 3 preimages.

A potentially large number of preimages (up to $2^N$) and the inability to use the parallel
scheme for the multiple mappings form Sec.~\ref{sec:parallel_evolvents} are the disadvantages
of the non-univalent evolvent.

@@

The methods of constructing the evolvents considered in the previous paragraphs produce the
curve $y(x)$, which in not a smooth one (see Fig.~\ref{fig:shifted_ev}). The absence of
smoothness may affect the properties of the reduced one-dimensional function $\varphi(y(x))$
adversely since a smooth curve reflect the information on the growth/decay of the initial
function better. On the basis of initial algorithm of constructing the non-smooth evolvent, a
generalized algorithm allowing constructing a smooth space-filling curve has been
proposed~\cite{Goryachih2017}. As an illustration, a smooth evolvent for the two-dimensional
case is presented in Fig.~\ref{fig:smooth}.
An increased computational complexity (several times as compared to the piecewise-linear
curves) is a disadvantage of the smooth evolvent. This caused by computing of the nonlinear smooth
functions. At that, the number of smooth intervals and the difficulty of the
computations of the curve increases with increasing accuracy of approximation and
space dimension.

%\end{Russian}

\begin{figure}[ht]
    \centering
    \subfloat[Smooth
evolvent]{{\includegraphics[width=.6\textwidth]{pictures/smooth.pdf}}\label{fig:smooth}}
    \subfloat[Numerical non-univalent
evolvent]{{\includegraphics[width=.4\textwidth]{pictures/noninjective.pdf}}\label{fig:noninje
ctive}}
    \caption{Different evolvents}
\end{figure}

@@

%------------------------------------------------------------------------------
\subsection{Comparison of the sequential evolvents}
\label{sec:seq_comp}
%\begin{Russian}
In order to understand whether any type of evolvents listed above has an essential advantage as
compared to other ones, the operating characteristics of the index method with different types
of evolvents have been obtained for the classes GKLS 2d Simple and GKLS 3d Simple. The
global minimum was considered to be found if the algorithm generates a trial point $y^k$ in the
$\delta$-vicinity of the global minimizer, i.e. $\left\|y^k-y^\ast\right\|_\infty\leq\delta$. The size
of the vicinity was selected as $\delta = 0.01\left\|b-a\right\|_\infty$. In case of GKLS
$\delta=0.01$.

In all experiments, the evolvent construction density parameter $m=12$. The minimum value of
the reliability parameter \(r\) was found for each type of evolvents by scanning over a uniform grid
with the step \(0.1\).

On the GKLS 2d Simple class at the minimum \(r\), the non-univalent evolvent and the smooth
one provide a faster convergence (Fig.~\ref{fig:gkls2d_opt}). The same was observed at
\(r=5.0\) as well (Fig.~\ref{fig:gkls2d_acc}). In the latter case, the shifted evolvent and the
rotating one begin to lag behind the rest since the value \(r=5.0\) is too big for them.
\begin{figure}[ht]
    \centering

\subfloat[$r=5.0$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_same_r_opt_pt_op.pdf}}\label{fig:gkls2d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_opt_pt_op.pdf}}\label{fig:gkls2
d_opt}}
    \caption{Operating characteristics on GKLS 2d Simple class}
\end{figure}

On the GKLS 2d Simple class at the minimum \(r\), the non-univalent evolvents and multiple
ones have a considerable advantage over the single evolvent (Fig.~\ref{fig:gkls3d_opt}).
The value \(r=4.5\) is too big for the rotated evolvents and for the shifted one
(Fig.~\ref{fig:gkls3d_acc}).

\begin{figure}[ht]
    \centering

\subfloat[$r=4.5$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_same_r_opt_pt_op.pdf}}\label{fig:gkls3d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_opt_pt_op.pdf}}\label{fig:gkls3
d_opt}}
    \caption{Operating characteristics on GKLS 3d Simple class}
\end{figure}

\paragraph{Overhead costs when using the shifted evolvents.}
In all experiments presented above, the number of computations of the objective function from
the GKLS class was taken into account when plotting the operating characteristics. However, in
the case of the shifted evolvent, the index method solves the problem with the constraint \(g_0\)
from (\ref{6_g0}). At the points where \(g_0\) is violated, the value of the objective function is
not computed. Nevertheless, these points are stored in the search information producing the
additional computational costs. In Table~\ref{tab:shifted_g0}, the averaged numbers of calls to
\(g_0\) and to the objective function are presented. At \(L=3\), the constraint \(g_0\) was
computed almost 20 times more than the objective function \(\varphi\) i. e. \(95\%\) of the whole
search information account for the auxiliary points. Such overhead costs are acceptable when
solving the problems of small dimension with the computation costly objective functions.
However, when increasing dimensionality and total number of trials other types of evolvents are
preferred.

\begin{table}
\begin{center}
\caption{Averaged number of computations of \(g_0\) and of \(\varphi\) when solving the
problems from GKLS 3d Simple class using the shifted evolvent}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
  $L$ & $calc(g_0)$ & $calc(\varphi)$ & $\frac{calc(g_0)}{calc(\varphi)}$ ratio \\
  \hline
  2 & 96247.9  & 6840.14 & 14.07\\
  \hline
  3 & 153131.0 & 7702.82 & 19.88\\
  \hline
  \end{tabular}
  \label{tab:shifted_g0}
\end{center}
\end{table}

%------------------------------------------------------------------------------
\subsection{Parallel rotated evolvents}
\label{sec:results_parallel}
In order to evaluate the efficiency of the parallel algorithm from
Sec.~\ref{sec:parallel_evolvents}, the numerical experiments on the GKLS 4d (Hard, Simple)
classes and on the GKLS 5d (Hard, Simple) ones were conducted. The value of \(r\) in all
experiments was equal to 5.0, the size of the \(\delta\)-vicinity of the known solution was
increased up to 0.3. When solving the series of problems, up to 8 cluster nodes and up to 32
computational threads on each node were employed.

In Table~\ref{tab:iterations}, an averaged number of iterations when solving 100 problems from
each considered class is presented.
The number of iterations is reduced considerably with increasing the number of nodes and the
number of threads on each node (except the GKLS 4d Simple class at the transition from 1
node to 4 ones in the single thread mode).

@@

If one assumes the costs of parallelization to be negligible as compared to the costs of
computing the objective functions in the optimization problems, the speedup in time due to the
use of the parallel method would be equal to the speedup with respect to the number of
iterations. However, actually this suggestion is not always true. In all numerical experiment the
time of computing the objective function was approximately $10^{-3}$ sec. In
Table~\ref{tab:speedup}, the speedups in iterations and in time (in the pendent brackets) are
presented. In the first row of the table corresponding to the sequential mode, the averaged time
of solving a single problem is presented in the pendent brackets. One can see from the table that
for the GKLS 4d classes it is more efficient to utilize a single node in the multithread mode
whereas for solving more complex five-dimensional problems, the use of several nodes is better,
each node is operating in the parallel mode.
